---
title: "Peak Bloom Prediction Demo"
author: "Eager Learner"
date: "01/01/2022"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```

## Instructions

In this analysis we demonstrate a very simple way of predicting the peak bloom date in the coming decade for all four locations required by the competition.
The models used here are very simple and are using only the historic data for these four locations, but no other information or covariates.
At the end of this document ((#appendix-rnoaa)[Appendix A]), we also demonstrate a simple way to get historic temperature data for the four locations via the `rnoaa` package.


```{r}
library(tidyverse)
library(caret)
```


## Loading the data

The data for each of the three main sites is provided as simple text file in CSV format.
Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site.

The six columns in each data file are

* _location_ a human-readable location identifier (`string`).
* _lat_ (approximate) latitude of the cherry trees (`double`).
* _long_ (approximate) longitude of the cherry trees (`double`).
* _alt_ (approximate) altitude of the cherry trees (`double`).
* _year_ year of the observation (`integer`).
* *bloom_date* date of peak bloom of the cherry trees (ISO 8601 date `string`). The "peak bloom date" may be defined differently for different sites
* *bloom_doy* days since January 1st of the year until peak bloom (`integer`). January 1st corresponds to `1`.

In R, the data files can be read with `read.csv()` and concatenated with the `bind_rows()` function:

```{r}
cherry <- read.csv("data/washingtondc.csv") %>% 
  bind_rows(read.csv("data/liestal.csv")) %>% 
  bind_rows(read.csv("data/kyoto.csv"))
```


## Visualizing the time series


```{r, fig.width=8, fig.height=3, out.width='100%', fig.cap="Time series of peak bloom of cherry trees since 1880 at three different sites."}
cherry %>% 
  filter(year >= 1880) %>%
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_point() +
  geom_step(linetype = 'dotted', color = 'gray50') +
  scale_x_continuous(breaks = seq(1880, 2020, by = 20)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```



```{r}
library(rnoaa)
```

The list of stations can be retrieved using the `ghcnd_stations()` function. Note that the closest weather station to each city with continuously collected maximum temperatures are USC00186350 (Washington D.C.), GME00127786 (Liestal), JA000047759 (Kyoto), and CA001108395 (Vancouver).

```{r, eval = FALSE}
stations <- ghcnd_stations()
```

As a simple demonstration, we retrieve the average seasonal maximum daily temperature (in 1/10 °C) from these stations using our own `get_temperature()` function, which wraps the `ghcnd_search()` function in the `rnoaa` package. (N.b. `ghcnd_search()` returns a list. Each element of the list corresponds to an element of the `var` argument.)

```{r}
#' Get the annual average maximum temperature at the given station,
#' separated into the 4 meteorological seasons (Winter, Spring, Summer, Fall).
#' 
#' The seasons are span 3 months each.
#' Winter is from December to February, Spring from March to May,
#' Summer from June to August, and Fall from September to November.
#' Note that December is counted towards the Winter of the next year, i.e.,
#' temperatures in December 2020 are accounted for in Winter 2021.
#' 
#' @param stationid the `rnoaa` station id (see [ghcnd_stations()])
#' @return a data frame with columns
#'   - `year` ... the year of the observations
#'   - `season` ... the season (Winter, Spring, Summer, Fall)
#'   - `tmax_avg` ... average maximum temperate in tenth degree Celsius
get_temperature <- function (stationid) {
  ghcnd_search(stationid = stationid, var = c("tmax"), 
               date_min = "1950-01-01", date_max = "2022-01-31")[[1]] %>%
  mutate(year = as.integer(format(date, "%Y")),
         month = as.integer(strftime(date, '%m')) %% 12, # make December "0"
         season = cut(month, breaks = c(0, 2, 5, 8, 11),
                      include.lowest = TRUE,
                      labels = c("Winter", "Spring", "Summer", "Fall")),
         year = if_else(month == 0, year + 1L, year)) %>%
  group_by(year, season) %>%
  summarize(tmax_avg = mean(tmax, na.rm = TRUE))
}

historic_temperatures <-
  tibble(location = "washingtondc", get_temperature("USC00186350")) %>%
  bind_rows(tibble(location = "liestal", get_temperature("GME00127786"))) %>%
  bind_rows(tibble(location = "kyoto", get_temperature("JA000047759"))) %>%
  bind_rows(tibble(location = "vancouver", get_temperature("CA001108395"))) %>%
  pivot_wider(names_from = season, names_prefix = "temp_", values_from = tmax_avg)

```

We then predict the peak bloom date in two steps. Step 1 extrapolates average winter and spring temperature until 2031 using simple linear regression. Step 2 predicts the peak bloom date given the extrapolated temperatures, again using simple linear regression. 

```{r}
library(gbm)
library(dismo)

vancouver_cov <- filter(historic_temperatures, location == "vancouver") %>%
  drop_na()%>%
  mutate(lat = 49.2237, long = -123.1636)

cherry_temp <- left_join(cherry, historic_temperatures, by = c("location", "year")) %>%
  drop_na() %>%
  dplyr::select( year, lat, long, temp_Winter, temp_Spring,temp_Summer, temp_Fall, bloom_doy)

trControl <- trainControl(method = "cv", number = 10)

tuneGrid <- expand.grid(interaction.depth = c(1,3,5),n.trees = (5:10)*100, shrinkage = c(.1,.01), n.minobsinnode =10)

fullMod <- train(x = cherry_temp[,1:7],
      y = cherry_temp$bloom_doy,
      distribution = "gaussian",
      method="gbm",
      metric = "RMSE",
      tuneGrid = tuneGrid,
      trControl = trControl)
plot(predict(mod, newData= vancouver_cov))

summary(mod)

```

```{r, variable selection}
cherry_test <- cherry_temp %>%
  filter(year == 2020)

cherry_train <- cherry_temp %>%
  filter(year <2020)

fullmod <- train(x = cherry_train[,1:7],
      y = cherry_train$bloom_doy,
      distribution = "gaussian",
      method="gbm",
      metric = "RMSE",
      tuneGrid = tuneGrid,
      trControl = trControl)

summary(mod)

predictions <- predict(fullmod, cherry_test)
RMSE_Full <- sqrt(mean((cherry_test$bloom_doy - predictions)^2))
  

RMSE <- c()
for(i in 1:ncol(cherry_test)-1){
  vars <- 1:ncol(cherry_test)-1
  vars <- vars[-i]
  mod <- train(x = cherry_train[,vars],
      y = cherry_train$bloom_doy,
      distribution = "gaussian",
      method="gbm",
      metric = "RMSE",
      tuneGrid = tuneGrid,
      trControl = trControl)
  predictions <- predict(mod, cherry_test)
  RMSE[i] <- sqrt(mean((cherry_test$bloom_doy - predictions)^2))
  
}


```