---
title: "Predictions Data"
author: "Grant Foster"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
source("FeatureFunctions.R")
library(tidyverse)
library(magrittr)
library(rnoaa)
```

```{r}
future_covs <- read.table(file="../data/covariates/future_climate.txt")
```

Assign a "cherry year" to match winter weather with spring blooms. 
```{r}
future_covs <- dplyr::filter(future_covs, month <5 | month >8)
future_covs$cherry_year <- 0

future_covs$cherry_year[future_covs$month >8] <- future_covs$year[future_covs$month >8]+1
future_covs$cherry_year[future_covs$month <8] <- future_covs$year[future_covs$month <8]
```

C_day as read in is unique integer identifier, where 1=Jan first. Below I reset to where September 1st is equal to 1, so that the climate dates relevant for a given bloom year are 1-243. 

```{r}
future_covs$C_day <- future_covs$C_day-243 #Set Sept1-Dec31 as 1-122
future_covs$C_day[future_covs$C_day<0] <- future_covs$C_day[future_covs$C_day<0] - min(future_covs$C_day) + max(future_covs$C_day)  
```

Joining in 2021 winter data so we can predict 2022. 

```{r}
historic_covs <- read.table(file="../data/covariates/historic_climate.txt")

future_covs <- dplyr::filter(historic_covs, year>=2021 & month >8) %>% mutate(., C_day=C_day-243, cherry_year=year+1) %>% rbind(., future_covs) #Add in 2021 Winter

#As of right now, we only have spring dates from liestal and kyoto; I remove the predictions to take the actual data from historic
future_covs <- dplyr::filter(future_covs, cherry_year!=2022 | location!="kyoto" | month != 1)
future_covs <- dplyr::filter(future_covs, cherry_year!=2022 | location!="kyoto" | month != 2 | day >15)

future_covs <- dplyr::filter(future_covs, cherry_year!=2022 | location!="vancouver" | month != 1)
future_covs <- dplyr::filter(future_covs, cherry_year!=2022 | location!="vancouver" | month != 2 | day >15)

future_covs <- dplyr::filter(historic_covs, year==2022) %>% mutate(., C_day=C_day+122, cherry_year=year) %>% rbind(., future_covs) #Add in 2022 Spring observations for Vancouver and Kyoto up to Feb 15th 

temp <- future_covs
```


The stations we used for Washington DC and Liestal are both missing the most recent few months of data. The code below fills in those gaps with other, close-by stations. The replacement station for washington is <5km away from the original, and the replacement station for Liestal is <15km away from the original. 
```{r}
get_historic <- function (stationid, location, var) {
  res<-vector(mode = 'list', length=length(stationid))
  for(i in 1:length(stationid)){
    aa<-as.data.frame(ghcnd_search(stationid = stationid[i], var = var,
                                   date_min = "1950-01-01", date_max = "2022-02-15")[[1]])
    aa['year']<-substr(x=aa$date, start = 1, stop = 4)
    aa['month']<-as.numeric(substr(x=aa$date, start = 6, stop = 7))
    aa['day']<-substr(x=aa$date, start = 9, stop = 10)
    aa[which(aa$month==12),'month']<-0
    aa['season']<-cut(as.numeric(aa$month), breaks = c(0, 2, 5, 8, 11),
                      include.lowest = TRUE,
                      labels = c("Winter", "Spring", "Summer", "Fall"))
    aa[which(aa$month==0),'month']<-12
    aa['C_day']<-strftime(aa$date, format = '%j')
    aa['location']<-location[i]
    res[[i]]<-aa
  }
  res<-do.call(rbind,res)
  res<-res[,-c(4:6)]
}

rep_stationid<-c('USW00013743','SZ000001940')
location<-c('washingtondc','liestal')
historic_tmax<-get_historic(rep_stationid, location, var='tmax')
historic_tmin<-get_historic(rep_stationid, location, var='tmin')
historic_prcp<-get_historic(rep_stationid, location, var='prcp')
historic_climate_alt<-historic_tmax
historic_climate_alt['tmin']<-historic_tmin[match(paste(historic_climate_alt$location, historic_climate_alt$date),
                                              paste(historic_tmin$location, historic_tmin$date)),'tmin']
historic_climate_alt['prcp']<-historic_prcp[match(paste(historic_climate_alt$location, historic_climate_alt$date),
                                              paste(historic_prcp$location, historic_prcp$date)),'prcp']
historic_climate_alt<-historic_climate_alt[,-c(1,3)]

historic_climate_alt<- dplyr::filter(historic_climate_alt, year > 2020)
historic_climate_alt <- dplyr::filter(historic_climate_alt, location!= "washingtondc" | month > 10 | year > 2021) #Remove everything but the correct washington series

historic_climate_alt <- dplyr::filter(historic_climate_alt, location!= "liestal" | year > 2021) #Remove everything but the correct liestal series


historic_climate_alt$tmax <- historic_climate_alt$tmax/10
historic_climate_alt$tmin <- historic_climate_alt$tmin/10
historic_climate_alt$year <- as.numeric(historic_climate_alt$year)
historic_climate_alt$cherry_year <- 2022 #All our missing data is relevant to the 2022 prediction year

historic_climate_alt$C_day <- as.numeric(historic_climate_alt$C_day)

historic_climate_alt$C_day[historic_climate_alt$month > 7] <- historic_climate_alt$C_day[historic_climate_alt$month > 7]-243

historic_climate_alt$C_day[historic_climate_alt$month < 7] <- historic_climate_alt$C_day[historic_climate_alt$month < 7]+122
```

Here we cut out the predicted values for those sites and replace them with the close site.s 

```{r}
future_covs <- temp
future_covs <- dplyr::filter(future_covs, location != "washingtondc" | C_day < 62 | C_day > 168| cherry_year > 2022) #Take out all Washington dates pre Feb 16th, 2022

future_covs <- dplyr::filter(future_covs, location != "liestal" | C_day > 168 | C_day <123 | cherry_year > 2022)  #Take out prediction dates for Liestal

future_covs <- rbind(future_covs, historic_climate_alt)
```


Create features


```{r}
cherryPreds <- expand(future_covs, location, cherry_year)
colnames(cherryPreds) <- c("location", "year")
cherryPreds <- dplyr::filter(cherryPreds, year!=2033)

## Above
### Total Time Series
cherryPreds <- countDays(tmax>0, data=future_covs, columnName = "TmaxDaysAbove0") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>5, data=future_covs, columnName = "TmaxDaysAbove5") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>10, data=future_covs, columnName = "TmaxDaysAbove10") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>20, data=future_covs, columnName = "TmaxDaysAbove20") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>25, data=future_covs, columnName = "TmaxDaysAbove25") %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- countDays(tmin>0, data=future_covs, columnName = "TminDaysAbove0") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmin>5, data=future_covs, columnName = "TminDaysAbove5") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmin>10, data=future_covs, columnName = "TminDaysAbove10") %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- countDays(prcp>0, data=future_covs, columnName = "PrcpDays") %>% left_join(cherryPreds, ., by=c("year", "location")) 

### Months
cherryPreds <- countDays(tmax>0, data=future_covs, columnName = "TmaxDecAbove0", month=12) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>0, data=future_covs, columnName = "TmaxJanAbove0", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax>0, data=future_covs, columnName = "TmaxFebAbove0", month=12) %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- countDays(prcp>0, data=future_covs, columnName = "DecPrcpDays", month=12) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(prcp>0, data=future_covs, columnName = "JanPrcpDays", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(prcp>0, data=future_covs, columnName = "FebPrcpDays", month=2) %>% left_join(cherryPreds, ., by=c("year", "location")) 


## Below
cherryPreds <- countDays(tmax<0, data=future_covs, columnName = "TmaxDaysSub0") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax<5, data=future_covs, columnName = "TmaxDaysSub5") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax<10, data=future_covs, columnName = "TmaxDaysSub10") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax<20, data=future_covs, columnName = "TmaxDaysSub20") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmax<25, data=future_covs, columnName = "TmaxDaysSub25") %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- countDays(tmin<0, data=future_covs, columnName = "TminDaysSub0") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmin<5, data=future_covs, columnName = "TminDaysSub5") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmin<8, data=future_covs, columnName = "TminDaysSub8") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- countDays(tmin<10, data=future_covs, columnName = "TminDaysSub10") %>% left_join(cherryPreds, ., by=c("year", "location")) 

#First Day Thresholds
#Full Time Series
cherryPreds <- firstDay(tmax<7, data=future_covs, columnName = "FirstTmaxSub7") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- firstDay(tmin<0, data=future_covs, columnName = "FirstTminFreeze") %>% left_join(cherryPreds, ., by=c("year", "location")) 


### Months
cherryPreds <- firstDay(tmax>4, data=future_covs, columnName = "FirstJanTmaxAbove4", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- firstDay(tmax>5, data=future_covs, columnName = "FirstFebTmaxAbove5", month=2) %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- firstDay(tmin>(-1), data=future_covs, columnName = "FirstJanTminAboveNeg1", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- firstDay(tmin>(-2), data=future_covs, columnName = "FirstFebTminAboveNeg2", month=2) %>% left_join(cherryPreds, ., by=c("year", "location")) 


#Max consec Thresholds
cherryPreds <- consecDays(tmin<0, data=future_covs, columnName = "MaxConsecTminSub0") %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- consecDays(tmax<0, data=future_covs, columnName = "MaxConsecTmaxSub0") %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- consecDays(tmin<0, data=future_covs, columnName = "MaxOctConsecTminSub0", month=10) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- consecDays(tmin<0, data=future_covs, columnName = "MaxNovConsecTminSub0", month=11) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- consecDays(tmin<0, data=future_covs, columnName = "MaxDecConsecTminSub0", month=12) %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- consecDays(tmin>0, data=future_covs, columnName = "MaxJanConsecTminAbove0", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- consecDays(tmin>0, data=future_covs, columnName = "MaxFebConsecTminAbove0", month=2) %>% left_join(cherryPreds, ., by=c("year", "location")) 

cherryPreds <- consecDays(tmax>10, data=future_covs, columnName = "MaxJanConsecTmaxAbove10", month=1) %>% left_join(cherryPreds, ., by=c("year", "location")) 
cherryPreds <- consecDays(tmax>10, data=future_covs, columnName = "MaxFebConsecTmaxAbove10", month=2) %>% left_join(cherryPreds, ., by=c("year", "location")) 

```

NA Interpolation. 
```{r}
table(is.na(cherryPreds))

for(i in 8:ncol(cherryPreds)){
  for(j in 1:length(unique(cherryPreds$location)))
  cherryPreds[which(cherryPreds$location==unique(cherryPreds$location)[j]),i] <- zoo::na.approx(cherryPreds[which(cherryPreds$location==unique(cherryPreds$location)[j]),i], rule=2)
}
table(is.na(cherryPreds))
```

```{r eval=FALSE}
#write.csv(cherryPreds, file = "PredictionCovariates.csv", row.names = FALSE)
```

Look for close-by stations (how we found alternative stations used above)
```{r}
DC_latlong <- data.frame("latitude"=38.88535, "longitude"=-77.038628, "id"="DC")
DC_Stations <- rnoaa::meteo_nearby_stations(DC_latlong, year_min = 2021, year_max=2022)

Lie_latlong <- data.frame("latitude"=47.4814, "longitude"=	7.730519, "id"="DC")
Lie_Stations <- rnoaa::meteo_nearby_stations(Lie_latlong, year_min = 2021, year_max=2022)
```



